---
title: "Parameter Optimization"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    self_contained: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=F, warning = F)
```

# Background

Consenus ICA is performed by running ICA on a matrix of normalized and scaled expression scores for
top variable genes. Each consensus ICA replicate consists of a number of subreplicates, 
whose components are clustered to yield consensus modules.

This process is replicated independently several times for each choice of `k`.


```{r load-pkgs}
library(tidyverse)
```

```{r load-metrics}
enr.supp <- read_csv(snakemake@input[['enr_supp']])
#enr.supp <- read_csv("~/work/ica_community_grid_search/test/enr/enr-metrics.csv")

enr <- read_csv(snakemake@input[['enr']])
#enr <- read_csv("~/work/ica_community_grid_search/test/enr/enr.csv")

sil <- read_csv(snakemake@input[['sil']])
#sil <- read_csv("~/work/ica_community_grid_search/test/ica/silhouette.csv")

dists <- read_csv(snakemake@input[['dists']])
#dists <- read_csv("~/work/ica_community_grid_search/test/ica/dists.csv")

knn <- snakemake@params[['knn']]
max_dist <- snakemake@params[['max_dist']]
```

```{r organize-inputs}
enr.supp <- enr.supp %>%
  mutate_at(vars('cov','pct_unique_term','pct_enr'), scales::rescale) %>%
  mutate(score = cov * pct_unique_term)

enr <- enr %>%
  filter(Significant > 2)
```

# Distances

For each component identified in all replicates of a given trial 
(trial: set of sub-replicates at a given k), `r knn` nearest neighbors were found
and the mean of those distances to the original component represents the outlier 
score for each component. Inspection of the the distribution of outlier scores for a given `k`
may yield a reasonable choice for cutoff.

```{r outlier-distrib}
dists %>%
  ggplot(aes(outlier.score,as.factor(comps))) +
  stat_density(aes(fill=after_stat(density)),geom='raster', position='identity', interpolate=F) +
  scale_fill_viridis_c(direction = 1, option = 'B') +
  geom_vline(xintercept = max_dist, color='white', linetype='dashed') +
  theme_minimal() +
  ylab("k") +
  facet_wrap(~rep) +
  theme(aspect.ratio = 0.5, plot.caption = element_text(hjust=0)) +
  labs(caption='Outlier score distribution by replicate for each k.')
```

# Silhouette

*After filtering as described above* silhouette scores were calculated for each component at a given k across replicates. This score was then averaged for each consensus component in each overall cICA replicate.

```{r mean-sil}
sil %>% group_by(cluster,comps,rep) %>%
  summarize(silhouette = mean(sil_width)) %>%
  group_by(comps,rep) %>%
  summarize(silhouette = mean(silhouette)) %>%
ggplot(aes(comps,silhouette)) +
  stat_summary(geom='line', color='red', linetype='dotted') +
  #geom_jitter(width = 0.1) +
  stat_summary(aes(comps,silhouette, color=silhouette)) +
  theme_classic() +
  xlab("k") +
  theme(aspect.ratio = 0.5, plot.caption = element_text(hjust=0)) +
  labs(caption='Silhouette scores across cICA replicates at a given k.')
```

# Unique enrichment x Coverage

Unique enrichment score is calculated by finding the percentage of the top 5 enriched terms for each
consensus module that are unique to that module.

Coverage represents the percentage of terms in a given GO ontology that are
represented within the top 5 significant enriched terms for each consensus component.

```{r, cov-uniq}
enr.supp %>%
  dplyr::rename(coverage='cov', uniqueness='pct_unique_term') %>%
  gather(metric, value, -rep:-comps) %>%
  filter(metric %in% c("coverage","uniqueness")) %>%
ggplot() +
  geom_smooth(aes(comps, value, color=metric)) +
  facet_wrap(~rep) +
  theme_classic() +
  scale_color_viridis_d(direction=-1, ) +
  theme(aspect.ratio = 0.5, plot.caption = element_text(hjust=0)) +
  xlab("k") + ylab('score') +
  labs(caption = "Relationship between k and enrichment metrics.")
```

These scores are scaled to lie between 0 and 1 and then multiplied to yield a single score
amenable for jointly optimizing `k` and FDR cutoffs.

```{r go-score}
ggplot(enr.supp,aes(as.factor(comps),as.factor(qval),fill=score)) +
  geom_raster(interpolate = T) +
  theme_minimal() +
  xlab("k") + ylab("qval") +
  scale_fill_viridis_c() +
  theme(aspect.ratio = 1, plot.caption = element_text(hjust=0)) +
  labs(caption = "Joint score displayed by FDR cutoff and k.")
```

```{r}
optimal <- enr.supp %>% arrange(desc(score)) %>% head(1) %>% dplyr::select(qval,comps) %>%
  gather(metric,value) %>%
  deframe()
```

The optimal `k` is `r optimal[['comps']]` and optimal q value cutoff is `r optimal[['qval']]`.

# Module sizes

The distributions of post-cutoff module sizes are displayed for each overall cICA replicate.

```{r mod-sizes}
fls <- Sys.glob(paste0("../test/ica/",optimal['comps'],"/*/consensus-ica.qvalues.csv.gz"))
tst <- map_df(fls,read_csv, .id = 'rep')

tst %>% filter(qval < optimal['qval']) %>% group_by(module,rep) %>% tally() %>%
  ggplot(aes(n)) +
  geom_histogram(aes(fill=..count..), color='black') +
  theme_classic() +
  scale_y_continuous(expand = c(0,0)) +
  facet_wrap(~rep) +
  scale_fill_fermenter(direction = 1, palette = 13) +
  theme(aspect.ratio = 0.5, plot.caption = element_text(hjust=0)) +
  xlab('module size') +
  labs(caption = 'Distribution of module sizes for each replicate at optimal params.')
```

# TE counts in enriched modules

The counts of TE features in each top TE-containing module are displayed by replicate.

```{r top-te-cts}
top_te <- tst %>% 
  filter(qval < optimal['qval']) %>%
  group_by(module, rep) %>%
  summarize(n_te = sum(!str_detect(X1,"FBgn"))) %>%
  arrange(desc(n_te)) %>%
  split(.,.$rep) %>% map(pull,module)

map(top_te,~head(.,3)) %>%
  imap_dfr(~filter(tst, module %in% .x & rep == .y)) %>%
  filter(qval < optimal['qval']) %>%
  group_by(rep,module) %>%
  summarize(n.te = sum(!str_detect(X1,"FBgn")), 
            n.gene = sum(str_detect(X1,"FBgn"))) %>%
  gather(feature, n, -rep,-module) %>%
  ggplot(aes(as.factor(module),n, fill=feature)) +
  geom_col() +
  facet_wrap(~rep, scales='free') +
  theme_classic() +
  xlab("module") +
  scale_fill_brewer(palette = 7) +
  scale_y_continuous(expand=c(0,0)) +
  theme(aspect.ratio = 1, plot.caption = element_text(hjust=0)) +
  labs(caption = "Modules yielding most TEs for each replicate at optimal k and FDR.")
```


# TE module enrichment in optimal choice

The top GO term enrichments for the Top 2 TE-containing modules in each overall replicate are displayed.

```{r fig.width=10}

map(top_te,~head(.,2)) %>%
  imap_dfr(~filter(enr, cluster %in% .x & rep == .y)) %>%
  filter(comps == optimal['comps']) %>%
  filter(qval == optimal['qval']) %>%
  group_by(rep, cluster) %>%
  top_n(10,score) %>%
  ggplot(aes(reorder(Term,score),score)) +
  facet_wrap(rep~cluster, scales = 'free') +
  geom_col(aes(fill=as.factor(rep):as.factor(cluster)), color='black') +
  coord_flip() +
  theme_classic() +
  scale_fill_brewer(type = 'qual', name="rep:module") +
  theme(aspect.ratio = 1, plot.caption = element_text(hjust=0)) +
  ylab("-log10(pval)") + xlab('Term')  +
  labs(caption = 'Top GO enrichment for Top TE-containing cluster found at optimal parameters.')
```


